Sparse spectral methods for partial differential equations on spherical caps

### 2021-04-15 papers/spherical-caps/revised/response-to-referees.tex

\item \comment{Is the code used to produce the numerical experiments going to be made publicly available? For example in the ApproxFun Julia package?}

The package that makes up the framework for expanding functions, obtaining the operator matrices etc. is purely experimental at this stage, but is publicly available. This fact and where it can be found has been added to the paper.

\end{itemize}

\centerline{\textbf{Other Changes}}

\begin{itemize}[parsep=1em,leftmargin=1em]
	\item Typo has been corrected in one equation in Section 2.3 (missing z from $Q(x,y,z)$).
	\item Proof for quadrature rule has been fixed - reference to $\bm{s}_l$ removed; $(1-x^2)^{-1/2}$ weight needed adding to the integrals over $dx$; clarified the maths that shows the symmetry that is referenced.
\end{itemize}

### 2021-04-15 papers/spherical-caps/revised/response-to-referees.tex

\centerline{\textbf{Response to Referee 1}}

\begin{itemize}[parsep=1em,leftmargin=1em]

\item \comment{The authors spend an inadaquate amount of time discussing other approaches for solving this problem and comparing their methodology (e.g., closest point method, radial basis functions). What is the complexity convergence behavior of other approaches that can handle spherical caps?}

We have added in a discussion of the closest point method. We have chosen to omit discussion on radial basis functions as we do not feel there is particularly suitable to reference and appropriate to discuss.

### 2021-04-15 papers/spherical-caps/revised/spherical-caps.tex
Changed to quad rule proof to fix some typos and make clearer
Addressed comments on CPM/RBF and DCT

@@

\begin{theorem}\label{Theorem:quadrule}
Let $M_1, M_2 \in \N$ and denote the $M_1$ Gauss quadrature nodes and weights on $[\alpha,1]$ with weight $(t - \alpha)^a$ as $(t_j, w_j^{(t)})$. Further, denote the $M_2$ Gauss quadrature nodes and weights $[-1,1]$ with weight $(1 - x^2)^{-\half}$ as $(s_j, w_j^{(s)})$.
Define for $j = 1,\dots,M_1, \: l=1,\dots,M_2$:
\begin{align*}
	\big(x_{l+(j-1)M_2}, \: y_{l+(j-1)M_2} \big) &:= \rho(t_j) \: \mathbf{s}_l, \\
	z_{l+(j-1)M_2} &:= t_j, \\
	w_{l+(j-1)M_2} &:= w_j^{(t)} w_l^{(s)}.
\end{align*}
Let $f(x,y,z)$ be a function on $\Omega$, and $N \in \N$. The quadrature rule is then
\begin{align*}
	\int_\Omega f(x,y,z) \: \genjacw^{(a,0)}(z) \: \D A \approx \sum_{j=1}^{M} w_j \: \big[ f(x_j, y_j, z_j) + f(-x_j, -y_j, z_j) \big],
\end{align*}
where $M = M_1 \: M_2$, and the quadrature rule is exact if $f(x,y,z)$ is a polynomial of degree $\le N$ with $M_1 \ge \half (N+1), M_2 \ge N+1$.
\end{theorem}

@@

Firstly, we note that
\begin{align*}
	\int_0^{2\pi} g\big(\cos \theta, \sin\theta\big) \: \D \theta = \int_{-1}^1 \Big( g\big(x, \sqrt{1-x^2}\big) + g\big(x, -\sqrt{1-x^2}\big) \Big) \frac{\D x}{\sqrt{1-x^2}}
\end{align*}
for some function $g$. Then, integrating the even function $f_e$ we have
\begin{align*}
	&\int_\Omega f_e(x,y,z) \: \genjacw^{(a,0)}(z) \: \D A \\
	&\quad \quad = \int_\alpha^1 \genjacw^{(a,0)}(z) \: \Big( \int_0^{2\pi} f_e\big(\rho(z)\cos\theta,\rho(z)\sin\theta, z\big) \: \D \theta \Big) \: \D z \\
	&\quad \quad = 2 \int_\alpha^1 \genjacw^{(a,0)}(z) \: \Big( \int_{-1}^1 f_e\big(\rho(z)x,\rho(z)\sqrt{1-x^2}, z\big) \: \D x \Big) \: \D z \\
	&\quad \quad \approx \int_\alpha^1 \genjacw^{(a,0)}(z) \: \Big( \sum_{l=1}^{M_2} w_l^{(s)} f_e\big(\rho(z) s_l, \rho(z) \sqrt{1 - s_l^2}, z\big) \Big) \: \D z \quad (\star) \\
	&\quad \quad \approx \sum_{j=1}^{M_1} w_j^{(t)} \sum_{l=1}^{M_2} w_l^{(s)} f_e\big(\rho(t_j) s_l, \rho(t_j) \sqrt{1 - s_l^2}, t_j\big) \quad (\star \star) \\
	&\quad \quad = \sum_{k=1}^{M_1 M_2}  w_j \: f_e(x_j, y_j, z_j).
\end{align*}
Suppose $f$ is a polynomial in $x,y,z$ of degree $N$, and hence that $f_e$ is a degree $\le N$ polynomial. It follows that $s \mapsto f_e\big(\rho(z)s,\rho(z)\sqrt{1-s^2}, z\big)$ for fixed $z$ is then a polynomial of degree $\le N$. We therefore achieve equality at $(\star)$ if $2M_2 - 1 \ge N$ and we achieve equality at $(\star \star)$ if also $2M_1 - 1 \ge N$.

Integrating the odd function $f_o$ results in
\begin{align*}
	&\int_\Omega f_o(x,y,z) \: \genjacw^{(a,0)}(z) \: \D A \\
	&\quad \quad = \int_\alpha^1 \genjacw^{(a,0)}(z) \: \Big( \int_0^{2\pi} f_o\big(\rho(z)\cos\theta,\rho(z)\sin\theta, z)\big) \: \D \theta \Big) \: \D z \\
	&\quad \quad = \int_\alpha^1 \genjacw^{(a,0)}(z) \: \Big( \int_{-1}^1 \Big[ f_o\big(\rho(z)x,\rho(z)\sqrt{1-x^2}, z\big) + f_o\big(\rho(z)x, - \rho(z)\sqrt{1-x^2}, z\big) \big] \: \D x \Big) \: \D z \\
	&\quad \quad = \int_\alpha^1 \genjacw^{(a,0)}(z) \: \Big( \int_{-1}^1 \Big[ f_o\big(\rho(z)x,\rho(z)\sqrt{1-x^2}, z\big) - f_o\big(\rho(z)x, \rho(z)\sqrt{1-x^2}, z\big) \big] \: \D x \Big) \: \D z \\
	&\quad \quad = 0.
\end{align*}
since $f_o(x,y,z) = - f_o(-x, -y, z)$. Hence, for a polynomial $f$ in $x,y,z$ of degree $N$,
\begin{align*}
	\int_\Omega f(x,y,z) \: \genjacw^{(a,0)}(z) \: \D A &= \int_\Omega \Big(f_e(x,y,z) + f_o(x,y,z)\Big) \:  \genjacw^{(a,0)}(z) \: \D A  \\
	&= \int_\Omega f_e(x,y,z) \: \genjacw^{(a,0)}(z) \: \D A \\
	&= \sum_{j=1}^{M}  w_j \: f_e(x_j, y_j, z_j),
\end{align*}
where $M = M_1 M_2$ and $2M_1 - 1 \ge N, 2M_2 - 1 \ge N$.
\end{proof}

@@

>>>>

There are of course other approaches for solving PDEs on surfaces, such as the closest point method \cite{macdonald2010implicit, macdonald2008level}. This involves solving the PDE inside a 3D volume that acts as a \enquote{shell} for the surface, and using a \enquote{closest point} to as the solution on the surface. While such an approach is useful for other geometries, it is of low order and does not achieve spectral convergence with sparse discretisations. Further, for our interest of the sphere, it would not take advantage of any rotational symmetry.

@@

\begin{theorem}\label{Theorem:quadrule}
	Let $M_1, M_2 \in \N$ and denote the $M_1$ Gauss quadrature nodes and weights on $[\alpha,1]$ with weight $(t - \alpha)^a$ as $(t_j, w_j^{(t)})$. Further, denote the $M_2$ Gauss quadrature nodes and weights $[-1,1]$ with weight $(1 - x^2)^{-\half}$ as $(s_j, w_j^{(s)})$.
Define for $j = 1,\dots,M_1, \: l=1,\dots,M_2$:
\begin{align*}
	x_{l+(j-1)M_2} &:= \rho(t_j) \: s_l, \\
	y_{l+(j-1)M_2} &:= \rho(t_j) \: \sqrt{1-s_l^2}, \\
	z_{l+(j-1)M_2} &:= t_j, \\
	w_{l+(j-1)M_2} &:= w_j^{(t)} w_l^{(s)}.
\end{align*}
Let $f(x,y,z)$ be a function on $\Omega$, and $N \in \N$. The quadrature rule is then
\begin{align*}
	\int_\Omega f(x,y,z) \: \genjacw^{(a,0)}(z) \: \D A \approx \sum_{j=1}^{M} w_j \: \big[ f(x_j, y_j, z_j) + f(-x_j, -y_j, z_j) \big],
\end{align*}
where $M = M_1 \: M_2$, and the quadrature rule is exact if $f(x,y,z)$ is a polynomial of degree $\le N$ with $M_1 \ge \half (N+1), M_2 \ge N+1$.
\end{theorem}

@@

Firstly, we note that
\begin{align*}
	\int_0^{2\pi} g\big(\cos \theta, \sin\theta\big) \: \D \theta &= \int_{-1}^1 \Big[ g\big(x, \sqrt{1-x^2}\big) + g\big(x, -\sqrt{1-x^2}\big) \Big] \frac{\D x}{\sqrt{1-x^2}} \\
	&= \int_{-1}^1 \Big[ g\big(x, \sqrt{1-x^2}\big) + g\big(-x, -\sqrt{1-x^2}\big) \Big] \frac{\D x}{\sqrt{1-x^2}}
\end{align*}
for any function $g$, using a change of variables $x \to -x$ for the second term in the integral. Then, integrating the even function $f_e$ we have
\begin{align*}
	&\int_\Omega f_e(x,y,z) \: \genjacw^{(a,0)}(z) \: \D A \\
	&\quad \quad = \int_\alpha^1 \genjacw^{(a,0)}(z) \: \Big( \int_0^{2\pi} f_e\big(\rho(z)\cos\theta,\rho(z)\sin\theta, z\big) \: \D \theta \Big) \: \D z \\
	&\quad \quad = \int_\alpha^1 \genjacw^{(a,0)}(z) \: \Big( \int_{-1}^1 \Big[ f_e\big(\rho(z)x,\rho(z)\sqrt{1-x^2}, z\big) \\
	&\quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad + f_e\big(-\rho(z)x, - \rho(z)\sqrt{1-x^2}, z\big) \Big] \: \frac{\D x}{\sqrt{1-x^2}} \Big) \: \D z \\
	&\quad \quad = 2 \int_\alpha^1 \genjacw^{(a,0)}(z) \: \Big( \int_{-1}^1 f_e\big(\rho(z)x,\rho(z)\sqrt{1-x^2}, z\big) \: \frac{\D x}{\sqrt{1-x^2}} \Big) \: \D z \\
	&\quad \quad \approx \int_\alpha^1 \genjacw^{(a,0)}(z) \: \Big( \sum_{l=1}^{M_2} w_l^{(s)} f_e\big(\rho(z) s_l, \rho(z) \sqrt{1 - s_l^2}, z\big) \Big) \: \D z \quad (\star) \\
	&\quad \quad \approx \sum_{j=1}^{M_1} w_j^{(t)} \sum_{l=1}^{M_2} w_l^{(s)} f_e\big(\rho(t_j) s_l, \rho(t_j) \sqrt{1 - s_l^2}, t_j\big) \quad (\star \star) \\
	&\quad \quad = \sum_{k=1}^{M_1 M_2}  w_j \: f_e(x_j, y_j, z_j).
\end{align*}
Suppose $f$ is a polynomial in $x,y,z$ of degree $N$, and hence that $f_e$ is a degree $\le N$ polynomial. It follows that $s \mapsto f_e\big(\rho(z)s,\rho(z)\sqrt{1-s^2}, z\big)$ for fixed $z$ is then a polynomial of degree $\le N$. We therefore achieve equality at $(\star)$ if $2M_2 - 1 \ge N$ and we achieve equality at $(\star \star)$ if also $2M_1 - 1 \ge N$.

Integrating the odd function $f_o$ results in
\begin{align*}
	&\int_\Omega f_o(x,y,z) \: \genjacw^{(a,0)}(z) \: \D A \\
	&\quad \quad = \int_\alpha^1 \genjacw^{(a,0)}(z) \: \Big( \int_0^{2\pi} f_o\big(\rho(z)\cos\theta,\rho(z)\sin\theta, z)\big) \: \D \theta \Big) \: \D z \\
	&\quad \quad = \int_\alpha^1 \genjacw^{(a,0)}(z) \: \Big( \int_{-1}^1 \Big[ f_o\big(\rho(z)x,\rho(z)\sqrt{1-x^2}, z\big) \\
	&\quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad + f_o\big(-\rho(z)x, - \rho(z)\sqrt{1-x^2}, z\big) \Big] \: \frac{\D x}{\sqrt{1-x^2}} \Big) \: \D z \\
	&\quad \quad = 0.
\end{align*}
since $f_o(x,y,z) = - f_o(-x, -y, z)$. Hence, for a polynomial $f$ in $x,y,z$ of degree $N$,
\begin{align*}
	\int_\Omega f(x,y,z) \: \genjacw^{(a,0)}(z) \: \D A &= \int_\Omega \Big(f_e(x,y,z) + f_o(x,y,z)\Big) \:  \genjacw^{(a,0)}(z) \: \D A  \\
	&= \int_\Omega f_e(x,y,z) \: \genjacw^{(a,0)}(z) \: \D A \\
	&= \sum_{j=1}^{M}  w_j \: f_e(x_j, y_j, z_j),
\end{align*}
where $M = M_1 M_2$ and $2M_1 - 1 \ge N, 2M_2 - 1 \ge N$.
\end{proof}

@@

\remark{While it may be possible to leverage a discrete cosine transform (DCT) to speed up the calculation of the coefficients  \cite{slevinsky2018use}, this is an open question.}
